---
title: "Example introduction: Cormack-Jolly-Seber model with dipper data"
subtitle: "April 2018"
author: "Perry de Valpine, UC Berkeley"
output:
  html_document:
    code_folding: show
    toc: yes
---

```{r setup, include=FALSE} 
library(methods) ## needed only when building documents outside of R
library(nimble)
if(!exists("nimble_course_dir")) ## set your own nimble_course_dir if needed
    nimble_course_dir <- file.path(getwd(),'..')
cur_dir <- getwd()
setwd(file.path(nimble_course_dir,
                 'examples_code',
                 'CJS_dipper'))
source("dipper_basic.R")
setwd(cur_dir)
```

# Dipper data

- Very common example
- Used for NIMBLE in Turek et al. (2016), Bayesian Analysis.
- We ignore covariates for sex and flood years or order to create the simplest possible model for software illustration.  (We don't recommend skipping important variables in real analyses!)
- 294 birds
- 7 years

# Standard BUGS code

Typical BUGS code for this model would be something like this:

```{r dipper-basic-model}
dipper_code
```

### Node definitions

- `phi` = survival probability
- `p` = detection (capture) probability
- `x[i, t]` = alive (1) or dead (0) for individual `i` at time `t`
- `y[i, t]` = captured (1) or not captured (0) for individual `i` at time `t`.
- `first[i]` = year of first capture for individual `i`

### Additional setup
- `x[i, t]` is data from first to last capture for individual `i`, and latent state thereafter.
- `y[i, t]` is not defined before first capture for individual `i`.
