---
title: "Introduction to using NIMBLE models"
subtitle: "April 2018"
author: "Perry de Valpine, UC Berkeley"
output:
  html_document:
    code_folding: show
    toc: yes
---

```{r setup, include=FALSE} 
library(methods) ## needed only when building documents outside of R
library(nimble)
if(!exists("nimble_course_dir")) ## set your own nimble_course_dir if needed
    nimble_course_dir <- file.path(getwd(),'..')
source(file.path(nimble_course_dir,
                 'examples_code',
                 'pump',
                 'pump_model.R'))
```

# Models are objects in R

# Get and set values

```{r, get-set-values}
pump$lambda
pump$lambda[3] <- 100
pump$lambda
```

# Simulate values

Simulation means:

* Draw values of stochastic nodes
* Calculate values of deterministic nodes

```{r, simulate}
pump$lambda
pump$simulate('lambda[1:5]')
pump$lambda
```

# Calculate values and log probabilities 

Calculation means:

* Calculate and store log probabilities of stochastic nodes
* Calculate values of deterministic nodes
* Return the sum of log probabilities of stochastic nodes

```{r, calculate}
pump$calculate(c('lambda[1:5]', 'theta[1:5]'))
```

# Query the graph structure

```{r, graph-query}
pump$getDependencies('lambda[1:5]')
```

# Model-generic programming

Say we want a function that simulates all parts of a model that depend on some input nodes and then returns the corresponding summed log probability.  I will call this part of the model "downstream".

```{r, generic-simulate}
simulate_downstream <- function(model, nodes) {
  downstream_nodes <- model$getDependencies(nodes, downstream = TRUE)
  model$simulate( downstream_nodes )
  logProb <- model$calculate( downstream_nodes )
  logProb
}
```
Notice that this function will work with *any* model and *any* set of input nodes.