---
title: "Introduction to compareMCMCs"
subtitle: "April 2018"
author: "Perry de Valpine, UC Berkeley"
output:
  html_document:
    code_folding: show
    toc: yes
    mathjax: deafult
---

```{r setup, include=FALSE} 
library(methods) ## needed only when building documents outside of R
library(nimble)
library(mcmcplots)
if(!exists("nimble_course_dir")) ## set your own nimble_course_dir if needed
    nimble_course_dir <- file.path(getwd(),'..')
cur_dir <- getwd()
setwd(file.path(nimble_course_dir,
                 'examples_code',
                 'CJS_dipper'))
source('dipper_model.R')
setwd(cur_dir)
```

# Automating MCMC comparisons in NIMBLE

- NIMBLE provides tools to automatically compare MCMC efficiency
across methods.
- Methods can include different NIMBLE MCMC configurations as well as
  JAGS, OpenBUGS/WinBUGS (not tested much) and Stan (separate code
  required).
- We will use `compareMCMCs`, which calls `MCMCsuite` internally.
- `MCMCsuite` can be used directy as well.

# Example: Metropolis-Hastings vs slice sampling. NIMBLE vs. JAGS

- `compareMCMCs` and `MCMCsuite` need to build and compile the model themselves.
- (This may become more flexible in the future.)
- These functions provide a variety of pre-defined MCMC configurations and a system
for providing your own.
- I will illustrate using slice samplers instead of the default
Metropolis-Hastings samplers for `p` and `phi`.
- (Assigning *all* samplers to be slice samplers is available as
  pre-defined option "nimble_slice" in the `MCMCs` argument.)

```{r, comparisons}
dipperInfo <- list(code=dipper_code,
                   constants=constants,
                   data=data, inits=inits)

## A function to replace default samplers with slice
## samplers for requested nodes
configureSlice <- function(Rmodel, nodes) {
    MCMCconf <- configureMCMC(Rmodel) ## get a default configuration
    ## print default samplers for illustration
    MCMCconf$printSamplers(c('p','phi'))
    for(node in nodes) {
        MCMCconf$removeSamplers(node)
        MCMCconf$addSampler(target = node, type = "slice")
    }
    ## print modified samplers for illustration
    MCMCconf$printSamplers(c('p','phi'))
    MCMCconf
}

## This is only needed during Rmarkdown evaluation to avoid Rmarkdown/NIMBLE non-standard evaluation conflicts
.GlobalEnv$configureSlice <- configureSlice

## A configuration list for compareMCMCs
MCMCdefs <- list(
    nimble_custom = quote({
        ## This code will be evaluated in an environment
        ## where 'Rmodel' is the model.
        customConf <- configureSlice(Rmodel, c('p', 'phi'))
        ## We must return a configuration
        customConf
    })
)

dipper_comparisons <- compareMCMCs(
    dipperInfo,
    MCMCs = c("jags", "nimble", "nimble_custom"),
    MCMCdefs = MCMCdefs,
    monitors = c("phi", "p"),
    niter = 10000,
    burnin = 1000,
    summary = FALSE)
```

```{r, comparison-results}
make_MCMC_comparison_pages(dipper_comparisons,
                           dir = "dipper_comparison_results",
                           modelNames = "dipper_comparisons")
```

The results are [here](dipper_comparison_results/dipper_comparisons.html)
