---
title: "Example: Zero-inflated Poisson N-mixture model for Swiss Great Tits data"
subtitle: "April 2018"
author: "Perry de Valpine, UC Berkeley"
output:
  html_document:
    code_folding: show
    toc: yes
---

```{r setup, include=FALSE} 
library(methods) ## needed only when building documents outside of R
library(nimble)
if(!exists("nimble_course_dir")) ## set your own nimble_course_dir if needed
    nimble_course_dir <- file.path(getwd(),'..')
```

# Example: Swiss Great Tits

This examples is from Applied Hierarchical Modeling in Ecology, Vol I (K&#233;ry and Royle), Chapter 6, section 6.11.1.

### Data summary:

- 267 1-km^2 quadrats in Switzerland
- 2-3 surveys per breeding season per quadrat
- Counts follow zero-inflated Poisson: There some probability of a structural zero, and otherwise a Poisson distribution.
- Latent abundance depends on 8 coefficients (intercept + 7 fixed effects)
- Detection probability depends on survey-specific intercept + 13 fixed effects (including some interactions).
- Random effects for site abundance, site detection and survey are not used in this example.

# Figure(s) from book

Let's run the original book code and generate its figure(s).

```{r make-figures}
cur_dir <- getwd()
setwd(file.path(nimble_course_dir,
                 'examples_code',
                 'ZIP_Nmixture_SwissGreatTits'))
.GlobalEnv$DO_PLOT <- TRUE
source("ZIP_Nmixture_SwissGreatTits.R")
setwd(cur_dir)
```

# Original BUGS code

Here is the original BUGS code, modified slightly for NIMBLE, including separation of posterior predictive quantities inside a definition-time conditional.

```{r original-code}
Section6p11_code
DO_POSTERIOR_PREDICTION <- FALSE
.GlobalEnv$DO_POSTERIOR_PREDICTION <- FALSE
```
