---
title: "Writing new distributions: multi-state capture-recapture"
subtitle: "April 2018"
author: "Perry de Valpine, UC Berkeley"
output:
  html_document:
    code_folding: show
    toc: yes
---

```{r setup, include=FALSE} 
library(methods) ## needed only when building documents outside of R
library(nimble)
if(!exists("nimble_course_dir")) ## set your own nimble_course_dir if needed
    nimble_course_dir <- file.path(getwd(),'..')
cur_dir <- getwd()
setwd(file.path(nimble_course_dir,
                 'examples_code',
                 'multi_state_CR_orchids'))
source("orchids_basic.R")
source("orchids_dDHMM.R")
setwd(cur_dir)

```

# Multi-state capture-recapture

- State/Capture history = [0 A 0 C B 0...] where "0" is not seen and A, B, C are observed states.

- When observed, states may be accurate or noisy.

- States may be individual condition (breeding/non-breeding), locations, or other.

- Model is defined by state transition and observation probabilities

# Orchids (ladyslipper) example

This example is described [here](../example_multi_state_CR_orchids/example_multi_state_CR_orchids_slides.html)

# BUGS code 

Let's get the BUGS code here for easy reference.

```{r basic-code}
orchids_code
```

- `ps[1, i, t, 2]` = probability of transition from state 1 to 2 for individual `i` at time `t`.
- `po[1, i, t, 2]` = probability of being observed in state 2 when in state 1 for individual `i` at time `t`.
- `s[t]` = survival probability at time `t`.
- `psiV[j]` = probability of transition from state 1 (Vegetative) to state j.
- `psiF[j]` = probability of transition from state 2 (Flowering) to state j.
- `psiD[j]` = probability of transition from state 3 (Dormant) to state j.

## Thinking about the code and the model's graph:

- Index `i` in `ps` and `po` is unnecessary in this model.   It creates thousands more edges (dependencies).
- Keeping indx `i` is ok for pretending we have a model with individual effects.
- According to Turek et al. 2016, this model has 2.157 latent variables.
- categorical distributions (`dcat`) are costly to sample with default methods: Samplers will try every possible value on every update, even when impossible or unlikely.

    - We could write a possibly more efficiency sampler in NIMBLE, but that's not our goal right now.

# Compare NIMBLE and JAGS on basic model

Let's run MCMC on the basic model.

```{r MCMC-basic}
parameters <- c("s", "psiV", "psiF", "psiD")
## One could argue to monitor the underlying random variables instead:
## parameters <- c('a', 'b', 'c', 's')


orchids_basic_comparison <- compareMCMCs(
    orchids_info,
    MCMCs = c("jags", "nimble"),
    monitors = parameters,
    niter = 20000,
    burnin = 2000,
    summary = FALSE
)

make_MCMC_comparison_pages(orchids_basic_comparison,
                           dir = "orchids_basic_comparison",
                           modelNames = "orchids_basic")

## browseURL(file.path("orchids_basic_comparison", "orchids_basic.html"))

```

Results are [here](orchids_basic_comparison/orchids_basic.html).

We could start working on customized sampler choices.

We can look more closely at the numbers like this:
```{r inspect-basic-MCMC}
orchids_basic_comparison[[1]]$summary
```

We see that `psiD`, especially `psiD[3]`, are the worst-mixing
parameters, so we could focus customization effort there.

But our goal here is to rewrite the model using multistate CR.

# Rewriting the model with a dynamic hidden Markov model distribution (dDHMM) #

We want the model to work like this

```{r orchids-DHMM-model}
orchids_DHMM_code
```

# Writing dDHMM

Here is a somewhat general version of a dynamic HMM distribution, from Turek et al. (2016).

```{r dDHMM}
dDHMMorchid <- nimbleFunction(
    run = function(x = double(1),    ## Observed capture (state) history
                   length = double(),## length of x (needed as a separate param for rDHMM)
                   prior = double(1),## 
                   Z = double(2),
                   T = double(3),
                   log = integer(0, default = 0)) {
        pi <- prior
        logL <- 0
        for(t in 1:length) {
            Zpi <- Z[x[t], ] * pi
            sumZpi <- sum(Zpi)
            logL <- logL + log(sumZpi)
            if(t != length)   pi <- (T[,,t] %*% asCol(Zpi) / sumZpi)[ ,1]
        }
        returnType(double())
        if(log) return(logL)
        return(exp(logL))
    }
)

rDHMMorchid <- nimbleFunction(
    run = function(n = integer(),
        length = double(), prior = double(1),
        Z = double(2),
        T = double(3)) {
        if(n != 1) print('should only specify n=1 in rDHMM() distribution')
        print('STILL NEED TO WRITE THE rDHMM() METHOD!')
        returnType(double(1))
        return(numeric(length, value = 1))
    }
)

```

### Some notes:

- The `prior` could be exchanged for a known initial state.
- Turek et al. set a 50/50 prior in Vegetative/Flowering, which I think ends up not mattering in this case.
- We call it "dynamic" because the transition matrix `T` can vary through time.
- One could make a narrower version that takes survival and transition vectors as inputs, rather than computing the `T` matrix in the model first.

# Run MCMC with the dDHMM version

- Slightly different inputs are needed.

```{r orchids-DHMM}
orchids_DHMM_comparison <- compareMCMCs(
    list(code = orchids_DHMM_code,
         data = orchids_DHMM_data,
         constants = orchids_DHMM_constants,
         inits = orchids_DHMM_inits()), 
    MCMCs = c("nimble"),
    monitors = parameters,
    niter = 20000,
    burnin = 2000,
    summary = FALSE
)

orchids_DHMM_comparison[[1]] <- rename_MCMC_comparison_method("nimble", "nimbleDHMM", orchids_DHMM_comparison[[1]])


orchids_all_comparison <- combine_MCMC_comparison_results(orchids_basic_comparison[[1]],
                                                         orchids_DHMM_comparison[[1]],
                                                         name = "orchids_MCMC_comparison")

make_MCMC_comparison_pages(orchids_all_comparison,
                           dir = "orchids_all_comparison",
                           modelNames = "orchids")

## browseURL(file.path("orchids_all_comparison", "orchids.html"))

```

Results are [here](orchids_all_comparison/orchids.html)

# Conclusions

- A good improvement was seen, but not as much as for the dipper example.
- One could still proceed with customizing samplers with the dDHMM version of the model.
- In the Canada Goose example, Turek et al. also grouped duplicate capture histories together.
