---
title: "Writing new distributions: Occupancy exercise"
subtitle: "April 2018"
author: "Perry de Valpine, UC Berkeley"
output:
  html_document:
    code_folding: show
    toc: yes
---

```{r setup, include=FALSE} 
library(methods) ## needed only when building documents outside of R
library(nimble)
if(!exists("nimble_course_dir")) ## set your own nimble_course_dir if needed
    nimble_course_dir <- file.path(getwd(),'..')
```

# An example occupancy model

We take our example from Applied Hierarchical Modeling in Ecology, Vol I (K&#233;ry and Royle): Chapter 10, Section 10.4.

- A simulated example
- `M` = 100 sites
- `J` = 3 visits to each site
- `vegHt` = vegetation height: logistic effect on occupancy probability
- `wind` = wind speed: logistic effect on detection probability

# Occupancy probabilities

Define `X[i, t]` = 1 if the species was detected as site `i` at time `t`, 0 otherwise.

Probability(`X[i, 1:T]`) = Probability(Site Occupied) Probability(Detections and non-detections) + Probability(Site Not Occupied) Indicator(All `X[i, t]` are 0)

# Let's see the figures from AHM book code

```{r AHM-figures}
cur_dir <- getwd()
setwd(file.path(nimble_course_dir,
                 'examples_code',
                 'occupancy_with_covariates'))
.GlobalEnv$DO_PLOT <- TRUE
source("occupancy_with_covariates.R")
setwd(cur_dir)
```

# BUGS code for this model

Here is the BUGS code provided by AHM.

- It may be slightly modified to work for NIMBLE.
- Derived (posterior predictive) quantities are shown here but omitted below when we run MCMC.

    - These quantities will not be available when we re-write the model.  Keeping them in the basic model would be an unfair computational cost in this comparison.

```{r AHM-BUGS-code}
Section10p4_code
```

## Thinking about the graph

- There is a latent state, `z[i]` for every site.

## Version we will use

The version without derived quantities is this:
```{r AHM-BUGS-noDerived-code}
Section10p4_noDerived_code
```

# Basic MCMC comparison

```{r basic-MCMC}
params_of_interest <- c("alpha0", "alpha1", "beta0", "beta1")
## N.occ would be of interest, but we'll need to get that after
## MCMC in the next version, so we skip it here too.

ni <- 25000
nb <- 2000

occupancy_basic_comparison <- compareMCMCs(
  modelInfo = list(
    code = Section10p4_noDerived_code,
    data = occupancy_data,
    inits = occupancy_inits()
  ),
  ## making the comparison for all latent states in params
  ## would be cumbersome
  monitors = params_of_interest,
  MCMCs = c('nimble', 'jags'),
  summary = FALSE,
  burnin = nb,
  niter = ni
)

make_MCMC_comparison_pages(occupancy_basic_comparison,
                           modelNames = "Section10p4",
                           dir = "occupancy_basic_comparison")

##browseURL(file.path(occupancy_basic_comparison, "Section10p4.html"))

```

Results are [here](occupancy_basic_comparison/Section10p4.html)

# Exercise

Rewrite this model to use a `dOccupancy` distribution for the vector `y[i, 1:J]` of observations from each size `i`.  This should allow you to remove the `z[i]` nodes from the model.  You will also need to write `dOccupancy` itself.

Bonus: Can you also rewrite the model to simplify the graph involving `p` and `phi`?
