---
title: "Writing new distributions: Cormack-Jolly-Seber capture-recapture"
subtitle: "April 2018"
author: "Perry de Valpine, UC Berkeley"
output:
  html_document:
    code_folding: show
    toc: yes
---

```{r setup, include=FALSE} 
library(methods) ## needed only when building documents outside of R
library(nimble)
if(!exists("nimble_course_dir")) ## set your own nimble_course_dir if needed
    nimble_course_dir <- file.path(getwd(),'..')
cur_dir <- getwd()
setwd(file.path(nimble_course_dir,
                 'examples_code',
                 'CJS_dipper'))
source("dipper_basic.R")
source("dipper_CJS.R")
setwd(cur_dir)

```

#  Basic capture-recapture model: CJS

Capture history: [1 0 0 1 0 0 0]

1 = captured (alive)
0 = not seen

### One way to look at the probabilities

P([1 0 0 1 0 0 0]) = P(seen) P(survives + not seen) P(survives + not seen) P(survives + seen) P(never seen again for 3 years)

P(never seen again for 3 years) = P(dead) + P(survives + not seen) * P(never seen again for 2 years)

P(never seen again for 2 years) = .... recursive reasoning.

### Another way to look at the probabilities (Hidden Markov Model)

P([1 0 0 1 0 0 0]) = P(1) P(0 | 1) P(0 | [1 0]) P(1 | [1 0 0]) * ...

The important part for us is the software flexibility, not the math.

# Example: Dipper data

The dipper data are introduced [here](../example_CJS_dipper/example_CJS_dipper_slides.html).

### Think about the graph and the sampling
- Every time `phi` or `p` is sampled, the entire model needs recalculation, so those are costly.
- Every `x[i, t]` that is a latent state (after last capture) needs to be sampled.
- The conditional distribution of `phi` and `p` changes with the states of `x[i, t]`.

# Run MCMCs on basic model

We will compare JAGS, NIMBLE with default samplers, and NIMBLE with slice samplers on `phi` and `p`.

```{r compare-MCMC-basic}
dipper_basic_comparison <- compareMCMCs(
    dipper_info,
    MCMCdefs = list(
        nim_slice_p_phi = quote({
            MCMCconf <- configureMCMC(Rmodel)
            MCMCconf$removeSamplers(c("p","phi"))
            MCMCconf$addSampler("p", type = "slice")
            MCMCconf$addSampler("phi", type = "slice")
            MCMCconf
        })),
    MCMCs = c("jags", "nimble", "nim_slice_p_phi"),
    monitors = c("phi", "p"),
    niter = 10000,
    burnin = 1000,
    summary = FALSE)

make_MCMC_comparison_pages(dipper_basic_comparison,
                           dir = "dipper_basic_comparison",
                           modelNames = "dipper_basic")

## Optional way to open browser directly from R
## browseURL(file.path("dipper_basic_comparison", "dipper_basic.html"))
```

Results are [here](dipper_basic_comparison/dipper_basic.html)

# Rewrite the model with a distribution for capture histories

We know how to calculate the probability of a capture history, so we'd
like to use that as a distribution.  We want this code to work:
 
```{r dipper-CJS}
dipper_code_CJS
```

# Write the distribution for capture histories

We write the `dCJS` as a nimbleFunction.  I will write it in the
state-space model format because this is more generalizable.

```{r dCJS}
dCJS <- nimbleFunction(
    run = function(x = double(1),    ## standard name for the "data"
                   survival = double(),
                   capture = double(),
                   l = double(),
                   log = integer(0, default = 0) ## required log argument
                   ) {
        probAliveGivenHistory <- 1
        ## logProbData will be the final answer
        logProbData <- 0
        if(l == 1) {  ## l==1 should not occur, but be defensive
            return(logProbData) 
        }
        for(t in 2:l) {
            ## probAlive is P(Alive(t) | x(1)...x(t-1))
            ## probAliveGivenHistory is (Alive(t-1) | x(1)...x(t-1))
            probAlive <- probAliveGivenHistory * survival
            if(x[t] == 1) {
                ## ProbThisObs = P(x(t) | x(1)...x(t-1))
                probThisObs <- probAlive * capture
                probAliveGivenHistory <- 1
            } else {
                probAliveNotSeen <- probAlive * (1-capture)
                probThisObs <- probAliveNotSeen + (1-probAlive)
                probAliveGivenHistory <- probAliveNotSeen / probThisObs
            }
            logProbData <- logProbData + log(probThisObs)
        }
        if(log) return(logProbData)
        return(exp(logProbData))
        returnType(double())
    }
)
```

We also need to write an "r" function, but if it will not be used, it
can be a dummy placeholder (even this may not always be needed).

```{r rCJS}
rCJS <- nimbleFunction(
    run = function(n = integer(),
                   survival = double(),
                   capture = double(),
                   l = double()) {
        returnType(double(1))
        return(numeric(l)) ## dummy behavior
    }
)
```

# Run MCMCs on the new model

```{r MCMC-CJS}

dipper_CJR_comparison <- compareMCMCs(
    dipper_info_CJS,
    MCMCdefs = list(
        nimCJS_slice_p_phi = quote({
            MCMCconf <- configureMCMC(Rmodel)
            MCMCconf$removeSamplers(c("p","phi"))
            MCMCconf$addSampler("p", type = "slice")
            MCMCconf$addSampler("phi", type = "slice")
            MCMCconf
        })),
    MCMCs = c("nimble", "nimCJS_slice_p_phi"),
    monitors = c("phi", "p"),
    niter = 10000,
    burnin = 1000,
    summary = FALSE)

dipper_CJR_comparison[[1]] <- rename_MCMC_comparison_method("nimble", "nimbleCJS", dipper_CJR_comparison[[1]])

dipper_all_comparison <- combine_MCMC_comparison_results(dipper_basic_comparison[[1]],
                                                         dipper_CJR_comparison[[1]],
                                                         name = "dipper_MCMC_comparison")

make_MCMC_comparison_pages(dipper_all_comparison,
                           dir = "dipper_all_comparison",
                           modelNames = "dipper")

## browseURL(file.path("dipper_all_comparison", "dipper.html"))

```

Results are [here](dipper_all_comparison/dipper.html)

# Conclusions

- Removing latent states by numerically integrating (summing) over
them can be beneficial, sometimes enormously so.

- One might argue the `dCJS` requires technical understanding but then
the final BUGS code is easier to write and understand.

- The `dCJS` could be extended for time-specific survival and
  detection probability.
