---
title: "Block sampling coefficients in Nmixture example"
subtitle: "April 2018"
author: "Perry de Valpine, UC Berkeley"
output:
  html_document:
    code_folding: show
    toc: yes
---

```{r setup, include=FALSE} 
library(methods) ## needed only when building documents outside of R
library(nimble)
library(mcmcplots)
if(!exists("nimble_course_dir")) ## set your own nimble_course_dir if needed
    nimble_course_dir <- file.path(getwd(),'..')
cur_dir <- getwd()
setwd(file.path(nimble_course_dir,
                 'examples_code',
                 'Nmixture_with_covariates'))
source("Nmixture_with_covariates.R")
setwd(cur_dir)
```

# Refresher on the Nmixture with covariates example

- Simulated example from AHM Chapter 6.4.

- 3 sampling occasions at each of 100 sites split among 3 habitat types.

- Detection covariate is `Xwind`.  Abundance covariate is `XvegHt`.

- What needs to be sampled in this model?

    - `alpha0[1]` ... `alpha0[3]`, detection intercepts at 3 sampling times
    - `alpha1[1]` ... `alpha1[3]`, detection slopes at 3 sampling times
    - `beta0[1]` ... `beta0[3]`, abundance intercepts at 3 sampling times
    - `beta1[1]` ... `beta1[3]`, abundance slopes at 3 sampling times
    - `N[1]` ... `N[100]`, latent population size at 100 sites

- There are some posterior predictive nodes.

# Look at the BUGS code

```{r BUGS-code}
Section6p4_code
```

# What are default samplers?

```{r, show-samplers}
MCMCconf <- configureMCMC(Nmix_model)
MCMCconf$printSamplers()
```

# Think about scalar samplers

- By default NIMBLE often uses adaptive random-walk
Metropolis-Hastings samplers (`type = 'RW'`) on scalar nodes.  These
are computationally fast but do not always yield a move.

- An alternative is slice sampling (`type = 'slice'`).  These always
yield a move but are computationally slower.

- JAGS uses slice samplers more extensively than NIMBLE's default, but
you can always customize NIMBLE's choices.

- JAGS does not use Metropolis-Hastings, IIUC.

# Think about the samplers: posterior correlations

- Coefficients in linear predictors generally yield correlated
posteriors, for which block sampling may help.

    - Adaptive random-walk Metropolis-Hastings (RWMH) block sampler (`type =
    'RW_block'`) is computationally fast but may or may not mix well.

    - One can use repeats of RWMH to push for more mixing.

    - Automated Factor Slice Sampler (AFSS) (`type = 'AF_slice'`) is
      computationally slower but should create better mixing.

# Think about shared computation in the graph

- Coefficients in linear predictors by definition share graph
dependencies (or "descendents" or "children").

- By sampling them jointly with one of the block sampling strategies,
we can save computations.

# Think about the latent states

- Sampling the discrete latent states common to population models
raises special challenges.

- By default, NIMBLE and JAGS (IIUC) use slice samplers for
Poisson-distributed variables.

- Later, we can try writing a new sampler for these latent states.

# Exercise: compare some sampling strategies

Use NIMBLE to compare the following strategies:

- JAGS
- NIMBLE with default samplers
- NIMBLE with slice samplers on all nodes
- NIMBLE with the following customization:

    - `AF_slice` samplers on each pair of coefficients that enter the
      same linear predictor (e.g. `alpha0[1]` and `alpha1[1]`,
      `beta0[1]` and `beta1[1]`, etc.).




