---
title: "Example: multi-state capture-recapture with ladyslipper orchids"
subtitle: "April 2018"
author: "Perry de Valpine, UC Berkeley"
output:
  html_document:
    code_folding: show
    toc: yes
---

```{r setup, include=FALSE} 
library(methods) ## needed only when building documents outside of R
library(nimble)
if(!exists("nimble_course_dir")) ## set your own nimble_course_dir if needed
    nimble_course_dir <- file.path(getwd(),'..')
cur_dir <- getwd()
setwd(file.path(nimble_course_dir,
                 'examples_code',
                 'multi_state_CR_orchids'))
source("orchids_basic.R")
source("orchids_dDHMM.R")
setwd(cur_dir)

```

# Orchids (ladyslipper) example

This example is described [here](../example_multi_state_CR_orchids/example_multi_state_CR_orchids_slides.html)

- From Bayesian Population Analysis Using WinBUGS (K&#233;ry and Schaub)
- Orchids modeled in states "vegetative", "flowering", "dormant", "dead"
- Observations: "seen vegetative", "seen flowering", "not seen".
- k = 11 annual observation periods 
- n = 250 unique plants
- n = 236 plants seen before the last observation time

This is a simple multi-state model because the state when not seen is unknown.

Obviously multi-state models can be more complicated than this.

# BUGS code 

This code is from BPA:

```{r basic-code}
orchids_code
```

- `ps[1, i, t, 2]` = probability of transition from state 1 to 2 for individual `i` at time `t`.
- `po[1, i, t, 2]` = probability of being observed in state 2 when in state 1 for individual `i` at time `t`.
- `s[t]` = survival probability at time `t`.
- `psiV[j]` = probability of transition from state 1 (Vegetative) to state j.
- `psiF[j]` = probability of transition from state 2 (Flowering) to state j.
- `psiD[j]` = probability of transition from state 3 (Dormant) to state j.

