---
title: "Understanding variables, nodes, lifted nodes, and graphs"
subtitle: "April 2018"
author: "Perry de Valpine, UC Berkeley"
output:
  html_document:
    code_folding: show
    toc: yes
---

```{r setup, include=FALSE} 
library(methods) ## needed only when building documents outside of R
library(nimble)
library(igraph)
if(!exists("nimble_course_dir")) ## set your own nimble_course_dir if needed
    nimble_course_dir <- file.path(getwd(),'..')
source(file.path(nimble_course_dir,
                 'examples_code',
                 'pump',
                 'pump_model.R'))
source(file.path(nimble_course_dir,
                 'general_code',
                 'drawGraph.R'))
```

# Outline

In this module, we'll learn some terms and concepts for models in NIMBLE:

- the distinction between a *variable* and a *node*;
- how to find variables and nodes in a model;
- some different types of nodes;
- what *lifted nodes* are;
- an introduction to *model-generic programming*.

We will continue to use the pump model to illustrate concepts because it is so simple.

# Pump model

```{r pump-model}
pumpCode
```

# Variables and Nodes

- A **variable** is an object that contains one or more nodes.
- A **node** is an object declared in a model statement.
- Every node is *stochastic* (declared with `~`) or * deterministic (declared with `<-`).

Example:

- `lambda` is a variable in the pump model.
- `lambda[3]` is a node in the pump model.

A node can be scalar or multivariate.

One of the main features of the BUGS language is that a variable can contain a variety of different kinds of nodes.

Example:

- We could have a variable `X` that is 10 $\times$ 10.
- `X[1:10, 1]` could be a stochastic node following a Dirichlet distribution.
- Each `X[1, 2:5]`...`X[10, 2:5]` could be a vector deterministic node from some calculation.
- Each `X[1,6]`...`X[9, 10]` could be a scalar stochastic node following a gamma distribution.
- `X[10, 6:10]` could be a stochastic node following a multivariate normal.

# Accessing values of variables and nodes

### Accessing variables

```{r use-variable}
pump$lambda ## accesses the variable lambda
pump$lambda[3:5] ## accesses elements of lambda
## which may or may not be one node or multiple nodes.
```

### Accessing nodes
```{r use-node}
pump[['lambda[3]']] ## accesses node lambda[3]
```

You can also assign values (`<-`) to variables and nodes.

Example: If you didn't provide a good choice of initial values when you created the model (via `inits` argument to `nimbleModel`), you are free to assign values to the model via its variables or nodes.

# Why does accessing nodes seem strange in R?

- `pump$lambda[3]` feels "natural" in R.
- `pump[['lambda[3]']]` feels "strange" in R.

    - Why?  The index is in a character string!
    - `pump[['lambda[i]']]` **does not work**. `i` will not be evaluated
    - `pump$lambda[i]` will of course work.

The reason for the node name system is model-generic programming.

- Sometimes a node will be `X[2, 3]` or `beta[4, 6, 1]`.
- These are all scalar nodes.
- We need a way to use nodes that does not depend on how many indices they have. Hence we use node names.
- We will see this more concretely later.

# Learning about the variable and nodes in your model.

We can see all the node names like this:
```{r see-nodes}
pump$getNodeNames()
```

We can see just the node names from one variable or part of a variable like this:
```{r expand-nodes}
pump$expandNodeNames('lambda')
pump$expandNodeNames('lambda[2:5]')
```

We can see variable names like this:
```{r see-variables}
pump$getVarNames()
```

We can get information about a variable like this:
```{r see-varInfo}
pump$getVarInfo('lambda')
pump$getVarInfo('theta')
```

# logProb variables

Every stochastic node automatically has an element of a "logProb" variable associated
with it. The log probability (density) of the variable will be stored in this element when it is *calculated*.

logProb variables are simply named as `logProb_[variable name]`.

We sometimes call these "logProb nodes" out of habit, but they are not
nodes in the graph.

Example:
```{r see-logProb}
pump$logProb_theta ## These were all initialized with the same value
pump$logProb_x
```

You don't generally need to use logProb variables directly.  Once we start
programming, you will instead use
```{r getLogProb}
pump$getLogProb('x[3]')
```
However it is useful to understand the existence of the logProb variables.


# Finding logProb variables

### Scalar nodes
For scalar stochastic nodes, the logProb node is usually obvious.

For `theta[3]`, the log probability value will be in
`logProb_theta[3]`.

### Multivariate nodes
For multivariate stochastic nodes, it is not as obvious.

If `X[3:5,  2]` follows a multivariate normal, its log probability
value will be in `logProb_X[3, 2]`.

You can find logProb variables like this:
```{r see-logProb-vars}
pump$getVarNames(includeLogProb = TRUE)
```

# Topological ordering of nodes and why it matters

Suppose we want to change the value of `theta[3]` and update all relevant
parts of the graph.

(The meaning of "relevant" depends on our purpose of changing `theta[3]`.
In this case imagine we are making a Metropolis-Hastings (MCMC)
proposal and need to calculate its acceptance probability.)

The nodes that need to be updated are shown in red.


```{r, deps-graph, fig.cap=""}
## function drawGraph is defined in the R file for this module
drawGraph(pump, colorBy = pump$getDependencies('theta[3]'))
```

Note that **order matters**.  If we calculated the log probability
density of `x[3]` *before* calculating `lambda[3]`, then `x[3]` would
use the *old* value of `lambda[3]`.  That would be a mistake.

A valid order for calculations is called **topologically sorted**.  Most
of NIMBLE's model-querying functions will return nodes in tologically sorted order.

# Types of nodes.

Some definitions:

- LHS = "left-hand-side" of a BUGS declaration
- RHS = "right-hand-side" of a BUGS declaration

Every node falls into one of following categories **based on
how it appears in BUGS code**:

- deterministic: declared on LHS  with `<-`
- stochastic: declared on LHS with `~`
- RHSonly: appears only on RHS of one or more declarations, never on
the LHS.

Every node *also* falls into one of the following categories **based on where
it is in the graph**:

- top: Has no stochastic nodes "above" it in the graph.
- latent: Has stochastic nodes "above" and "below" it in the graph.
- end: Has no stochastic nodes "below" it in the graph.

Sometimes people use ancestray terms:

- "above" = "parent" or "ancestor"
- "below" = "child" or "descendent"

Once a model is created, stochastic nodes can be labeled as **data**.

- Multiple copies of the same model can be created from the same BUGS code, with
possibly different data nodes.
- The data label is used by programs to decide what to do.
- Values of data nodes can be changed.
- Data labels should not be changed after an algorithm like MCMC
  has been built for a model.

# Finding out about nodes

## Stochastic, deterministic, and/or data
Here is a graph color-coded by whether nodes are stochastic,
deterministic, and/or labeled as data:

```{r stoch-determ-data-fig, fig.cap = ""}
drawGraph(pump, colorBy = "stochDetermData")
```

You can select groups of nodes by these types:

```{r}
## deterministic nodes
pump$getNodeNames(determOnly = TRUE)
## all stochastic nodes
pump$getNodeNames(stochOnly = TRUE, includeData = TRUE)
## non-data stochastic nodes
pump$getNodeNames(stochOnly = TRUE, includeData = FALSE)
```

## top, latend, and end

Here is a graph color-coded by whether nodes are top, latent, or end
nodes.
```{r top-latent-end-fig, fig.cap=""}
drawGraph(pump, 'topLatentEnd')
```

You can also select groups of nodes by these types:

```{r}
pump$getNodeNames(topOnly = TRUE)   #salmon
pump$getNodeNames(latentOnly = TRUE)  #cyan
pump$getNodeNames(endOnly = TRUE)    #plum
```

## Combined categories

Say you want to know what the stochastic, latent nodes are.

```{r stoch-latent-nodes}
pump$getNodeNames(stochOnly = TRUE, latentOnly = TRUE)
```

## RHSonly nodes

You also can query the model for RHSonly nodes:
```{r RHSonly}
pump$getNodeNames(includeRHSonly = TRUE)
```

In this case, there are no RHSonly nodes.  We might think `t` nodes
are RHSonly, but in fact since `t` was provided in `constants`, it is
no longer considered to be a variable in the model.  Its values have been **permanently** substituted where needed.

We could modify how we create the model as follows to see `t` nodes appear as
RHSonly.

```{r see-RHSonly}
## Move "t" from constants to data
pumpData2 <- c(pumpData, pumpConsts['t'])
pumpConsts2 <- pumpConsts['N']
pump2 <- nimbleModel(
    pumpCode,
    data = pumpData2,
    constants = pumpConsts2
)
pump2$getNodeNames()
pump2$getNodeNames(includeRHSonly = TRUE)
```

Now we see that `t` nodes have appeared when we ask to include `RHSonly` nodes.

# Topologically sorting nodes

If for some reason you have a collection of node names and need them
topologically sorted:

```{r}
pump$topologicallySortNodes(c('x[1:3]','beta','theta[1:3]'))
```

Often there is more than one valid topological sorting, but NIMBLE
chooses one that is usually intuitive and uses it for your model.

# Inspecting the model: Querying graph dependencies

There are several kinds of dependencies that may be needed in different circumstances.

- *Stochastic dependencies* (called the *Markov blanket* in
     computer science) terminate at stochastic nodes
     (inclusive). These are the dependencies needed for many kinds of MCMC samplers.

```{r}
pump$getDependencies('theta[3]')
```

- *Deterministic dependencies* terminate at stochastic nodes but
omit the stochastic nodes. These can be useful for implementing a particle filter.
```{r}
pump$getDependencies('theta[3]', determOnly = TRUE)
```

```{r, echo=FALSE, fig.cap=""}
drawGraph(pump, pump$getDependencies('theta[3]', determOnly = TRUE))
```

- Other ways to control the dependency query can be found by
```{r}
args(pump$getDependencies)
## or help(modelBaseClass)
```


# The importance of querying a model: lifted nodes

**You should never assume you know what nodes are in a model simply
because you wrote the BUGS code.**

NIMBLE implements some features by inserting its own nodes.  These are
called **lifted nodes**.

Let's look at the two main ways this happens:

```{r, fig.cap=""}
m1 <- nimbleModel(
    nimbleCode({
        tau ~ dunif(0, 100)
        x ~ dnorm(0, tau) #by default, tau is a precision
    }))
plot(m1$getGraph())
m1$getNodeNames()
```
The node `lifted_d1_over_sqrt_oPtau_cP` has been inserted between
`tau` and `x`.   The resulting model would equivalently have been
created by this BUGS code:

```{r, eval = FALSE}
nimbleCode({
    tau ~ dunif(0, 100)
    lifted_d1_over_sqrt_oPtau_cP <- 1/sqrt(tau)
    x ~ dnorm(0, sd = lifted_d1_over_sqrt_oPtau_cP) #by default, tau is a precision
}))
```

NIMBLE has *lifted* the calculation of standard deviation from
precision so that it is part of the model's graph.  Therefore *you
will make a mistake if you assume that the dependencies of `tau`
include only `x`*:

```{r, eval = FALSE}
m1$calculate(c('tau','x')) ## Wrong: the lifted node is being neglected
```

The correct way to do this is with model-generic programming:
```{r, eval = FALSE}
m1$calculate( m1$getDependencies('tau') )
```

The second common situation that creates lifted nodes is:
```{r, fig.cap=""}
m2 <- nimbleModel(
    nimbleCode({
        a ~ dnorm(0, 1)
        b ~ dnorm(a + 1, 1)
    }))
plot(m2$getGraph())
m2$getNodeNames()
```

In this case, NIMBLE lifted the expression `a+1` and created a node
for it.   Again the correct way to program is to query a model to
learn about its nodes.  Do not assume you know what the nodes are.

